Can We Formalise Type Theory Intrinsically without Any Compromise?

### 2025-12-03 tex/CPP26/main.tex
Three remaining issues to resolve (termination, strictification, and the notion of models)

\documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
\setcopyright{acmlicensed}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{XXXXXXX.XXXXXXX}
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation email}{June 03--05,
  2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/2018/06}

%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[utf8]{inputenc}
\usepackage{newunicodechar}
\usepackage{xspace}
\usepackage{xcolor}
\usepackage{mathtools}
\let\Bbbk\relax
\usepackage{amsmath,amssymb,mathbbol}
\let\Bbbk\relax
\usepackage{bbold}

%% Remove the following if there are no todo items.
\setlength {\marginparwidth }{2cm}
\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Macros %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\CA}{\textsc{Cubical Agda}\xspace}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Can Natural Models Simplify The Metatheory of Type Theory in Cubical Agda?}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Liang-Ting Chen}
\affiliation{%
  \institution{Institute of Information Science, Academia Sinica}
  \city{Taipei}
  \country{Taiwan}}
\email{liangtingchen@as.edu.tw}

\author{Fredrik Nordvall Forsberg}
\affiliation{%
  \institution{Department of Computer and Information Sciences, University of Strathclyde}
  \city{Glasgow}
  \country{United Kingdom}}

\author{Tzu-Chun Tsai}
\affiliation{%
  \institution{Institute of Information Science, Academia Sinica}
  \city{Taipei}
  \country{Taiwan}}
%\affiliation{%
%  \institution{Institute for Logic, Language and Computation, University of Amsterdam}
%  \city{Amsterdam}
%  \country{Netherlands}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}

  We formalise type theory in type theory, using Awodey's notion of \emph{natural model} of type theory.
  The initial natural model can be represented using quotient inductive-inductive-recursive types in the proof assistant \CA.
Using natural models leads to fewer instances of so-called ``transport hell'' in the syntax of type theory and its elimination rules.
To show that the approach is feasible, we formalise some meta-properties, including the standard % and logical predicate
interpretation, normalisation by evaluation for typed terms, and strictification constructions.
Since our formalisation is carried out using \CA's built-in support for quotient types, all our constructions compute.
However, the use of transports along equations often reappeared when developing more sophisticated metatheory.
Hence it is still a considerable struggle to develop the metatheory of type theory using welltyped terms with current proof assistant technology, and the effort is about the same using natural models or not.

%   This paper presents a formalisation of a core type theory and its metatheory within \CA.
% The type theory features $\Pi$-types, an inductive type of Booleans, and a universe.
% We formalise key meta-properties, including standard and logical predicate interpretations, canonicity, and normalisation by evaluation for typed terms.

% The type theory is defined intrinsically as a representable map of presheaves using quotient inductive-inductive-recursive types.
% Moreover, type theory and its elimination are defined with dependent identities without transports.
% This work is the first formalization of type theory that uses \CA's built-in support for quotient types, without using Licata's workaround and any rewriting rules, whereas its proofs of canonicity and normalisation are computational.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003790.10011740</concept_id>
       <concept_desc>Theory of computation~Type theory</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

@@

%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Proof Assistants, Formalisation, Cubical Agda, Quotient Inductive-Inductive-Recursive Type}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.

@@

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle
\bibliographystyle{ACM-Reference-Format}

\section{Introduction}
% FNF (Fri 5 Sep)

Transport hell = transports appear inside terms that you actually want to study, hence you need to do a lot of reasoning about transport rather than the actual objects of interest.

\section{Setting and metatheory}
% FNF (Sun 7 Sep)

Cubical Agda with UIP

QIIRTs

\section{Type theory as a natural model}
% LTC (Sun 7 Sep)

\subsection{Natural models}

\subsection{The Syntax of Type Theory as a Quotient Inductive-Inductive-Recursive Type}

QIIRT def of the syntax

\subsection{Recursion and Elimination Principles}

\subsection{Strictification}

\section{Metatheory}
% LTC (Tue 9 Sep)

\subsection{Standard Model}

% Argue for relative consistency?

\subsection{Normalisation by Evaluation}

% (for SC)

\section{Comparison with Other Approaches}
% FNF (Tue 9 Sep)

Compared to QIIT:

\begin{itemize}
\item Fewer transports in the syntax, but they tend to come back in concrete models
\item Strictification orthogonal
\end{itemize}

Compared to untyped version:
\begin{itemize}
\item Untyped version might still be easiest to work with, with current proof assistant technology
\end{itemize}

\section{Discussion}
% LTC (at first)

Practical considerations (eg NBE computes, support for interleaved mutual definitions)

Support from proof assistant (eg safety guarantees)

Easier in OTT-based proof assistant

>>>>

\documentclass[sigplan,screen,natbib,pbalance]{acmart}
\settopmatter{printfolios=true,printccs=false,printacmref=false}
\setcopyright{acmlicensed}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{XXXXXXX.XXXXXXX}
\acmConference[CPP '26]{Certified Programs and Proofs}{January 12--13, 2026}{Rennes, France}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/2018/06}

%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{newunicodechar}
\usepackage{xspace}
\usepackage{xcolor}
%\usepackage{natbib}
\setcitestyle{numbers,sort&compress}
\usepackage{mathtools}
\let\Bbbk\relax
\usepackage{amsmath,amssymb,mathbbol}
\usepackage{bbold}
\usepackage[inline]{enumitem}

%% Remove the following if there are no todo items.
\setlength {\marginparwidth }{2cm}
\usepackage[obeyFinal,textsize=footnotesize]{todonotes}
\usepackage[capitalise]{cleveref}
\newcommand{\LT}[2][]{\todo[inline,author={L-T},caption={},color={pink},#1]{#2}}
\newcommand{\FNF}[2][]{\todo[inline,author={Fred},caption={},#1]{#2}}

\usepackage{microtype}
\usepackage{newunicodechar}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Macros %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\CA}{\textrm{Cubical Agda}\xspace}
\newcommand{\Agda}{\textrm{Agda}\xspace}
\newcommand{\Set}{\mathbf{Set}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% lhs2tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\cons}[1]{\mathbf{#1}}
\newcommand{\iden}{\mathit}

\newcommand{\awa}[2]{\mathrlap{#2}\phantom{#1}} % as wide as

\definecolor{addition}{RGB}{204,255,216}
\definecolor{keyword}{RGB}{204,255,216}
\definecolor{identifier}{RGB}{204,255,216}
\newcommand{\highlight}[2]{\smash{\text{\colorbox{#1}{\kern-.1em\vphantom{\vrule height 1.2ex depth 0.1ex}\smash{\ensuremath{#2}}\kern-.1em}}}}

\let\Bbbk\relax
%% ODER: format ==         = "\mathrel{==}"
%% ODER: format /=         = "\neq "
%
%
\makeatletter
\@ifundefined{lhs2tex.lhs2tex.sty.read}%
  {\@namedef{lhs2tex.lhs2tex.sty.read}{}%
   \newcommand\SkipToFmtEnd{}%
   \newcommand\EndFmtInput{}%
   \long\def\SkipToFmtEnd#1\EndFmtInput{}%
  }\SkipToFmtEnd

\newcommand\ReadOnlyOnce[1]{\@ifundefined{#1}{\@namedef{#1}{}}\SkipToFmtEnd}
\usepackage{amstext}
\usepackage{amssymb}
\usepackage{stmaryrd}
\DeclareFontFamily{OT1}{cmtex}{}
\DeclareFontShape{OT1}{cmtex}{m}{n}
  {<5><6><7><8>cmtex8
   <9>cmtex9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmtex10}{}
\DeclareFontShape{OT1}{cmtex}{m}{it}
  {<-> ssub * cmtt/m/it}{}
\newcommand{\texfamily}{\fontfamily{cmtex}\selectfont}
\DeclareFontShape{OT1}{cmtt}{bx}{n}
  {<5><6><7><8>cmtt8
   <9>cmbtt9
   <10><10.95><12><14.4><17.28><20.74><24.88>cmbtt10}{}
\DeclareFontShape{OT1}{cmtex}{bx}{n}
  {<-> ssub * cmtt/bx/n}{}
\newcommand{\tex}[1]{\text{\texfamily#1}}	% NEU

\newcommand{\Sp}{\hskip.33334em\relax}

\newcommand{\Conid}[1]{\mathit{#1}}
\newcommand{\Varid}[1]{\mathit{#1}}
\newcommand{\anonymous}{\kern0.06em \vbox{\hrule\@width.5em}}
\newcommand{\plus}{\mathbin{+\!\!\!+}}
\newcommand{\bind}{\mathbin{>\!\!\!>\mkern-6.7mu=}}
\newcommand{\rbind}{\mathbin{=\mkern-6.7mu<\!\!\!<}}% suggested by Neil Mitchell
\newcommand{\sequ}{\mathbin{>\!\!\!>}}
\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}
\usepackage{polytable}

%mathindent has to be defined
\@ifundefined{mathindent}%
  {\newdimen\mathindent\mathindent\leftmargini}%
  {}%

\def\resethooks{%
  \global\let\SaveRestoreHook\empty
  \global\let\ColumnHook\empty}
\newcommand*{\savecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\savecolumns[#1]}}
\newcommand*{\restorecolumns}[1][default]%
  {\g@addto@macro\SaveRestoreHook{\restorecolumns[#1]}}
\newcommand*{\aligncolumn}[2]%
  {\g@addto@macro\ColumnHook{\column{#1}{#2}}}

\resethooks

\newcommand{\onelinecommentchars}{\quad-{}- }
\newcommand{\commentbeginchars}{\enskip\{-}
\newcommand{\commentendchars}{-\}\enskip}

\newcommand{\visiblecomments}{%
  \let\onelinecomment=\onelinecommentchars
  \let\commentbegin=\commentbeginchars
  \let\commentend=\commentendchars}

\newcommand{\invisiblecomments}{%
  \let\onelinecomment=\empty
  \let\commentbegin=\empty
  \let\commentend=\empty}

\visiblecomments

\newlength{\blanklineskip}
\setlength{\blanklineskip}{0.66084ex}

\newcommand{\hsindent}[1]{\quad}% default is fixed indentation
\let\hspre\empty
\let\hspost\empty
\newcommand{\NB}{\textbf{NB}}
\newcommand{\Todo}[1]{$\langle$\textbf{To do:}~#1$\rangle$}

\EndFmtInput
\makeatother
%
%
%
%
%
%
% This package provides two environments suitable to take the place
% of hscode, called "plainhscode" and "arrayhscode". 
%
% The plain environment surrounds each code block by vertical space,
% and it uses \abovedisplayskip and \belowdisplayskip to get spacing
% similar to formulas. Note that if these dimensions are changed,
% the spacing around displayed math formulas changes as well.
% All code is indented using \leftskip.
%
% Changed 19.08.2004 to reflect changes in colorcode. Should work with
% CodeGroup.sty.
%
\ReadOnlyOnce{polycode.fmt}%
\makeatletter

\newcommand{\hsnewpar}[1]%
  {{\parskip=0pt\parindent=0pt\par\vskip #1\noindent}}

% can be used, for instance, to redefine the code size, by setting the
% command to \small or something alike
\newcommand{\hscodestyle}{}

% The command \sethscode can be used to switch the code formatting
% behaviour by mapping the hscode environment in the subst directive
% to a new LaTeX environment.

\newcommand{\sethscode}[1]%
  {\expandafter\let\expandafter\hscode\csname #1\endcsname
   \expandafter\let\expandafter\endhscode\csname end#1\endcsname}

% "compatibility" mode restores the non-polycode.fmt layout.

\newenvironment{compathscode}%
  {\par\noindent
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed\)%
   \par\noindent
   \ignorespacesafterend}

\newcommand{\compaths}{\sethscode{compathscode}}

% "plain" mode is the proposed default.
% It should now work with \centering.
% This required some changes. The old version
% is still available for reference as oldplainhscode.

\newenvironment{plainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\hspre\(\let\hspost\)%
   \pboxed}%
  {\endpboxed%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newenvironment{oldplainhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

% Here, we make plainhscode the default environment.

\newcommand{\plainhs}{\sethscode{plainhscode}}
\newcommand{\oldplainhs}{\sethscode{oldplainhscode}}
\plainhs

% The arrayhscode is like plain, but makes use of polytable's
% parray environment which disallows page breaks in code blocks.

\newenvironment{arrayhscode}%
  {\hsnewpar\abovedisplayskip
   \advance\leftskip\mathindent
   \hscodestyle
   \let\\=\@normalcr
   \(\parray}%
  {\endparray\)%
   \hsnewpar\belowdisplayskip
   \ignorespacesafterend}

\newcommand{\arrayhs}{\sethscode{arrayhscode}}

% The mathhscode environment also makes use of polytable's parray 
% environment. It is supposed to be used only inside math mode 
% (I used it to typeset the type rules in my thesis).

\newenvironment{mathhscode}%
  {\parray}{\endparray}

\newcommand{\mathhs}{\sethscode{mathhscode}}

% texths is similar to mathhs, but works in text mode.

\newenvironment{texthscode}%
  {\(\parray}{\endparray\)}

\newcommand{\texths}{\sethscode{texthscode}}

% The framed environment places code in a framed box.

\def\codeframewidth{\arrayrulewidth}
\RequirePackage{calc}

\newenvironment{framedhscode}%
  {\parskip=\abovedisplayskip\par\noindent
   \hscodestyle
   \arrayrulewidth=\codeframewidth
   \tabular{@{}|p{\linewidth-2\arraycolsep-2\arrayrulewidth-2pt}|@{}}%
   \hline\framedhslinecorrect\\{-1.5ex}%
   \let\endoflinesave=\\
   \let\\=\@normalcr
   \(\pboxed}%
  {\endpboxed\)%
   \framedhslinecorrect\endoflinesave{.5ex}\hline
   \endtabular
   \parskip=\belowdisplayskip\par\noindent
   \ignorespacesafterend}

\newcommand{\framedhslinecorrect}[2]%
  {#1[#2]}

\newcommand{\framedhs}{\sethscode{framedhscode}}

% The inlinehscode environment is an experimental environment
% that can be used to typeset displayed code inline.

\newenvironment{inlinehscode}%
  {\(\def\column##1##2{}%
   \let\>\undefined\let\<\undefined\let\\\undefined
   \newcommand\>[1][]{}\newcommand\<[1][]{}\newcommand\\[1][]{}%
   \def\fromto##1##2##3{##3}%
   \def\nextline{}}{\) }%

\newcommand{\inlinehs}{\sethscode{inlinehscode}}

% The joincode environment is a separate environment that
% can be used to surround and thereby connect multiple code
% blocks.

\newenvironment{joincode}%
  {\let\orighscode=\hscode
   \let\origendhscode=\endhscode
   \def\endhscode{\def\hscode{\endgroup\def\@currenvir{hscode}\\}\begingroup}
   %\let\SaveRestoreHook=\empty
   %\let\ColumnHook=\empty
   %\let\resethooks=\empty
   \orighscode\def\hscode{\endgroup\def\@currenvir{hscode}}}%
  {\origendhscode
   \global\let\hscode=\orighscode
   \global\let\endhscode=\origendhscode}%

\makeatother
\EndFmtInput
%
%
\ReadOnlyOnce{agda.fmt}%

\RequirePackage[T1]{fontenc}
\RequirePackage[utf8]{inputenc}
\RequirePackage{amsfonts}
\RequirePackage{newunicodechar}

\providecommand\mathbbm{\mathbb}

% TODO: Define more of these ...
\DeclareUnicodeCharacter{737}{\textsuperscript{l}}
\DeclareUnicodeCharacter{8718}{\ensuremath{\blacksquare}}
\DeclareUnicodeCharacter{8759}{::}
\DeclareUnicodeCharacter{9669}{\ensuremath{\triangleleft}}
\DeclareUnicodeCharacter{8799}{\ensuremath{\stackrel{\scriptscriptstyle ?}{=}}}
\DeclareUnicodeCharacter{10214}{\ensuremath{\llbracket}}
\DeclareUnicodeCharacter{10215}{\ensuremath{\rrbracket}}
\DeclareUnicodeCharacter{27E6}{\ensuremath{\llbracket}}
\DeclareUnicodeCharacter{27E7}{\ensuremath{\rrbracket}}
\DeclareUnicodeCharacter{2200}{\ensuremath{\forall}}

\DeclareUnicodeCharacter{2294}{\ensuremath{\sqcup}}
\DeclareUnicodeCharacter{2080}{\ensuremath{_0}}
\DeclareUnicodeCharacter{2081}{\ensuremath{_1}}
\DeclareUnicodeCharacter{2082}{\ensuremath{_2}}
\DeclareUnicodeCharacter{2083}{\ensuremath{_3}}

\DeclareUnicodeCharacter{2115}{\ensuremath{\mathbb{N}}}
\DeclareUnicodeCharacter{2236}{:}
\DeclareUnicodeCharacter{2237}{\ensuremath{\mathrel{::}}}
\DeclareUnicodeCharacter{03A3}{\ensuremath{\Sigma}}
\DeclareUnicodeCharacter{03BB}{\ensuremath{\lambda}}
\DeclareUnicodeCharacter{039B}{\ensuremath{\Lambda}}
\DeclareUnicodeCharacter{03C0}{\ensuremath{\pi}}
\DeclareUnicodeCharacter{03C3}{\ensuremath{\sigma}}
\DeclareUnicodeCharacter{03C9}{\ensuremath{\omega}}
\newunicodechar{β}{\ensuremath{\beta}}
\newunicodechar{η}{\ensuremath{\eta}}
\newunicodechar{Γ}{\ensuremath{\Gamma}}
\newunicodechar{Δ}{\ensuremath{\Delta}}
\newunicodechar{Θ}{\ensuremath{\Theta}}
\newunicodechar{Ξ}{\ensuremath{\Xi}}
\newunicodechar{Π}{\ensuremath{\Pi}}
\newunicodechar{τ}{\ensuremath{\tau}}
\newunicodechar{γ}{\ensuremath{\gamma}}
\newunicodechar{δ}{\ensuremath{\delta}}
\newunicodechar{∅}{\ensuremath{\emptyset}}
\newunicodechar{≡}{\ensuremath{\equiv}}
\newunicodechar{∘}{\ensuremath{\circ}}
\newunicodechar{𝔹}{\ensuremath{\mathbb{B}}}
\newunicodechar{𝕓}{\ensuremath{\mathbb{b}}}
\newunicodechar{∎}{\ensuremath{\blacksquare}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{⋆}{\ensuremath{\star}}
\newunicodechar{⌜}{\ensuremath{\ulcorner}}
\newunicodechar{⌝}{\ensuremath{\urcorner}}
\newunicodechar{ʸ}{\ensuremath{{}^y}}
\newunicodechar{ⁿ}{\ensuremath{{}^n}}
\newunicodechar{ᴾ}{\ensuremath{{}^P}}

\DeclareUnicodeCharacter{2032}{\ensuremath{\prime}}
\DeclareUnicodeCharacter{2113}{\ensuremath{\ell}}
\DeclareUnicodeCharacter{2207}{\ensuremath{\nabla}}
\DeclareUnicodeCharacter{220B}{\ensuremath{\ni}}
\DeclareUnicodeCharacter{2264}{\ensuremath{\leq}}
\DeclareUnicodeCharacter{21D2}{\ensuremath{\Rightarrow}}
\DeclareUnicodeCharacter{22A2}{\ensuremath{\vdash}}
\DeclareUnicodeCharacter{22A4}{\ensuremath{\top}}
\DeclareUnicodeCharacter{22A5}{\ensuremath{\bot}}

\DeclareUnicodeCharacter{1D57}{\ensuremath{^t}}
\newunicodechar{∙}{\ensuremath{\cdot}}

% TODO: This is in general not a good idea.
\providecommand\textepsilon{$\epsilon$}
\providecommand\textmu{$\mu$}

%Actually, varsyms should not occur in Agda output.

% TODO: Make this configurable. IMHO, italics doesn't work well
% for Agda code.

\renewcommand\Varid[1]{\mathord{\textsf{#1}}}
\let\Conid\Varid
\newcommand\Keyword[1]{\textsf{\textbf{#1}}}
\EndFmtInput

\renewcommand\Varid[1]{\mathord{\textsf{#1}}}
%\let\Conid\Varid
%\newcommand\Keyword[1]{\textsf{\textbf{#1}}}

%% end of the preamble, start of the body of the document source.

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Can We Formalise Type Theory Intrinsically without Any Compromise?}
\subtitle{A Case Study in \CA}
%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Liang-Ting Chen}
\affiliation{%
  \institution{Academia Sinica}
  \city{Taipei}
  \country{Taiwan}}
\email{liangtingchen@as.edu.tw}

\author{Fredrik Nordvall Forsberg}
\affiliation{%
  \institution{University of Strathclyde}
  \city{Glasgow}
  \country{United Kingdom}}
\email{fredrik.nordvall-forsberg@strath.ac.uk}

\author{Tzu-Chun Tsai}
\affiliation{%
  \institution{Academia Sinica}
  \city{Taipei}
  \country{Taiwan}}
%\affiliation{%
%  \institution{University of Amsterdam}
%  \city{Amsterdam}
%  \country{Netherlands}}

@@

\begin{abstract}
  We present an intrinsic representation of type theory in the proof assistant \CA, inspired by Awodey’s natural models of type theory.
  The initial natural model is defined as quotient inductive-inductive-recursive types, leading us to a syntax accepted by \CA without using any transports, postulates, or custom rewrite rules.
We formalise some meta-properties such as the standard model, normalisation by evaluation for typed terms, and strictification constructions.
Since our formalisation is carried out using \CA's native support for quotient inductive types, all our constructions compute at a reasonable speed.
When we try to develop more sophisticated metatheory, however, the `transport hell' problem reappears.  Ultimately, it remains a considerable struggle to develop the metatheory of type theory using an intrinsic representation that lacks strict equations.
The effort required is about the same whether or not the notion of natural model is used.
\end{abstract}

%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
\begin{CCSXML}\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{ccs2012>}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
   <concept>
       <concept_id>10003752.10003790.10011740</concept_id>
       <concept_desc>Theory of computation~Type theory</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

@@

\keywords{Proof Assistants, Formalisation, Cubical Agda, Quotient Inductive-Inductive-Recursive Type}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.

@@

\maketitle
\bibliographystyle{ACM-Reference-Format}

\section{Introduction}

%\LT[inline]{Clearly articulate what has been formalised and what has not, when appropriate? Ideally finish formalisation, otherwise scale back and discuss what has been done. (R1, R3)}
%\LT[inline]{Alternative formulation? E.g., \url{https://web.archive.org/web/20241004151846/https://lists.chalmers.se/pipermail/agda/2019/011176.html} (R2)}

%\listoftodos

% FNF (Fri 5 Sep)

Internalising the syntax and semantics of type theory in type theory is a long-standing problem which stretches the limits of the theory~\cite{Dybjer1996,Danielsson2006,Chapman2009,McBride2010,Altenkirch2016a}.
%
There are both practical and theoretical reasons to pursue this problem.
%
On the practical side, an internal representation of type theory is needed for mechanised metatheory and metaprogramming.
%
On the theoretic side, if type theory is supposed to be a general constructive foundation of mathematics, then it should in particular be able to reason about its own syntax and semantics (up to inherent limitations due to G\"odel's incompleteness theorems).
%
In dependent type theory, types can depend on terms, which means that all of contexts, types and terms need to be defined simultaneously.
%
This is one reason why formalising type theory in type theory is hard.
%

Early approaches to formalising type theory, e.g.\ McKinna and Pollack~\cite{McKinna1999}, dealt with untyped terms that were later refined by a typing relation, setoid equality, or quotients, and thus had to prove a lot of congruence lemmas by hand~\cite{Danielsson2006,Chapman2009,Brunerie2019,Boer2020}.
%
A breakthrough was achieved by Altenkirch and Kaposi~\cite{Altenkirch2016a}, who showed that quotient inductive-inductive types (QIITs)~\cite{Altenkirch2018} can be employed to significantly simplify the internal representation of well typed terms (or, more precisely, derivations), since equality constructors can be used to represent equations such as $\beta$- and $\eta$-equality.
%
They took Dybjer's notion of a model of type theory in the form of a category with families~\cite{Dybjer1996}, and translated it into a QIIT definition.
%
In effect, this gives rise to the \emph{initial} category with families, with the elimination principles of the QIIT giving a unique morphism of categories with families to any other model.
%
This thus gives a both principled and practical way to formalise the syntax and equational theory of type theory in type theory; the feasibility of the approach was demonstrated by e.g.\ formalising normalisation by evaluation using this representation~\cite{Altenkirch2017}.

At the time of publication of Altenkirch and Kaposi~\cite{Altenkirch2016a}, the proof assistant \Agda did not allow equality constructors in data type declarations, so a workaround known as `Licata's trick'~\cite{Licata2011} was used by postulating (equality) constructors and writing down the eliminator explicitly, which meant giving up features of the proof assistant such as dependent pattern matching.
\CA, the cubical variant of \Agda~\cite{Vezzosi2021}, is now equipped with a native support for QIITs, so it is natural to ask if we can use this support to formalise the intrinsic representation of type theory without the trick or any other compromise.

In this paper, we explore this question and see what the proof assistant provides and lacks to achieve this goal.
We quickly find that their QIIT definitions break the strict positivity --- a syntactic restriction imposed by \CA to ensure consistency.
Moreover, their definition is cumbersome to work with, since the type of later constructors or even equations often only make sense because of earlier equations.
%
In an intensional type theory, such as those implemented in proof assistants, this manifests itself in transport terms across equality proofs \emph{inside} other terms, and leads to the so-called `transport hell' --- rather than just reasoning about the terms you actually want to study, you now also have to do a lot of reasoning about transports themselves and their algebraic properties.
%
It turns out that we need an alternative way of representing type theory intrinsically without any transport hell, in order to make our formalisation of type theory more lightweight and accepted by \CA.

%
The framework of categories with families is only one of several (more or less) equivalent notions of models of type theory~\cite{Hofmann1997}, and we were wondering if any of the other notions might offer any advantages.
%
Bense~\cite{Bense2024} suggested that Awodey's notion of natural model~\cite{Awodey2018} might be a good candidate.
%
Indeed, in a natural model, the indexing of terms over their types $\mathsf{Tm}_{\Gamma} : \mathsf{Ty}(\Gamma) \to \mathsf{Set}$ (as in a category with families) is replaced by a `fibred' perspective where each term instead \emph{has} a type, as picked out by a function $\mathsf{tyOf} : \mathsf{Tm}(\Gamma) \to \mathsf{Ty}(\Gamma)$.
%
Terms and types are still indexed by contexts $\Gamma$, but since most `type mismatches' arise from equations between types, not equations between contexts (indeed many formulations of type theory does not even have a notion of context equality), this should mean that many uses of transports can be avoided.

We test this hypothesis by formalising type theory in a form inspired by natural models.
%
\CA is particularly a good fit for such a project, because not only does it support QIITs, it also supports inductive-recursive types~\cite{Dybjer1999}, which are needed to simultaneously define the recursive $\mathsf{tyOf}$ function together with the inductively defined types $\mathsf{Tm}(\Gamma)$ and $\mathsf{Ty}(\Gamma)$.
%
Indeed, it could be the lack of support for inductive-recursive definitions in many proof assistants which has held back formalisation attempts based on natural models so far.

While we manage to avoid transports occurring in its own syntax, the experiment is not an outright success.
%
Indeed, we found that when developing more sophisticated metatheory, such as when defining a logical predicate model, the use of transports along equations often reappeared.
%
Strictification~\cite{Donko2022,Kaposi2025}, a recent technique for strictifying equations, cannot be applied directly because of the lack of strict proposition in the metatheory.
%
Furthermore, we found that the use of natural models is less well supported in the Cubical Agda of today, compared to approaches based purely on QIITs.
%
This is because we are more reliant on the computational behaviour of the recursively defined $\mathsf{tyOf}$ function, and this behaviour is only available in `later' clauses, which leads to the need for hacks and tricks to work around this limitation.
%
We discuss proof assistant features and their helpfulness further towards the end of the paper, after presenting our formalisation.

\paragraph{Contributions} We make the following contributions:
\begin{itemize}
\item We present an intrinsically well typed representation of the syntax of type theory, inspired by Awodey's natural models (\cref{sec:tt}).
\item We derive elimination and recursion principles for the syntax (\cref{sec:tt:elim}), and show how the standard model and the term model are constructed (\cref{sec:standard-model}).
\item We discuss strictification constructions on models, and show that they also apply to our notion of natural models (\cref{sec:strictify}).
\item We develop normalisation by evaluation for substitution calculus~(\cref{sec:nbe}) as a proof of concept: our development is carried out in \CA, which has a computational implementation of QIITs and principles such as function extensionality, so the resulting normaliser computes, and can potentially be extracted as a verified program.
\item We discuss pros and cons of our approach compared to other approaches, and which features of a proof assistant and its metatheory would make future formalisation more feasible (\cref{sec:discussion}).
\end{itemize}

%\LT{the idea of using natural model appears at least in 2024 \cite{Bense2024}, and it is a natural idea to formalise type theory in this way.}
\section{Setting and metatheory}
% FNF (Sun 7 Sep)

%\LT{Should we discharge this assumption in our formalisation, eg set-truncate where needed instead? (R1, R2)}
Our formalisation is carried out in \CA without the use of the \ensuremath{\Conid{Glue}} type and, in particular, the univalence principle, checked using the option \text{\ttfamily \char45{}\char45{}cubical\char61{}no\char45{}glue} available in the forthcoming \Agda 2.9.0.
%
We explicitly set-truncate the types we define.
Therefore, the term \emph{set} is used interchangeably with type.
%
For the sake of simplicity we occasionally postulate uniqueness of identity proofs (UIP) locally, which is of course inconsistent with univalence but in principle compatible with cubical type theory. 
%
We believe that the cubical type theory XTT~\cite{Sterling2022}, which enjoys definitional UIP without univalence, justifies this local assumption.
%
